CrypTool Script \\ Mathematics and Cryptography

### 2003-03-27 script/en/chapters/primes.tex
be updates
more be updates

The sequence of prime numbers starts with $$ 2,~ 3,~ 5,~ 7, ~ 11, ~ 13, ~
17, ~ 19, ~ 23, ~ 29, ~ 31, ~ 37, ~ 41, ~ 43, ~ 47, ~ 53, ~ 59, ~ 61,
~ 67, ~ 71, ~ 73, ~ 79, ~ 83, ~ 89, ~ 97, \cdots . $$
The first 100 numbers include precisely 25 prime numbers. After this,
the percentage of primes constantly decreases. Prime numbers can be
factorised in a uniquely {\em trivial} way: 
$$5 = 1 \cdot 5,\quad  17 = 1 \cdot 17, \quad 1,013 = 1 \cdot 1,013,  \quad
1,296,409 = 1 \cdot 1,296,409.$$
All numbers that have $2$ or more factors are called \index{Number!composite}
{\em composite}. These include $$ 4 = 2 \cdot 2, \quad 6 = 2\cdot 3 $$ as well
as numbers that {\em look like primes}, but are in fact composite:
$$ 91 = 7 \cdot 13, \quad 161=7 \cdot 23, \quad 767 =13 \cdot 59. $$

@@

\begin{theorem}\label{thm-pz-prod}
Each element $n$ of the natural numbers greater than $1$ can be written as the
product $n = p_1 \cdot p_2 \dots p_m$ of prime numbers. If two such
factorisations $$n =  p_1 \cdot p_2 \cdots p_m = p'_1 \cdot p'_2 \cdots
p'_{m'}$$ are given, then they can be reordered such that $\;m = m'\;$ and for
all $i$:  $\;p_i = p'_i$.
\end{theorem}

@@

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid%
\footnote{Euclid,
a Greek mathematician of 4th and 3rd century B.C. He worked at the
Egyptian academy of Alexandria and wrote ``The Elements'', the most well 
known systematically textbook of the Greek mathematics.} 
knew this in the third century before Christ.
\begin{theorem}[Euclid\footnote{The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. The theorem has already been distinguished
and proven in Euclidís Elements (Book IX, s. 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as followed
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than 
any previously existing amount of prime numbers.
}]\label{thm-pz-euklid} % Ende der Fuﬂnote
% Fussnote VOR "]" in [Euclid], damit kein Leerraum vor der Fussnotennummer.
% Vorher stand da: ...(Euclid). BLANK Fussnote und das Blank stoerte.
% Nun steht die Fussnote direkt hinter "Euclid" und vor der ")".
% Eigentlich h‰tte ich sie gerne direkt hinter "(Euclid)", noch vor dem 
% automatisch gesetzten Punkt. 
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}
His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (proof by contradiction). He assumed that there is only a
finite number of primes and therefore a largest prime number. Based on this
assumption, he drew logical conclusions until he obtained an obvious
contradiction. This meant that something must be wrong. As there were no
mistakes in the chain of conclusions, it could only be the assumption that was
wrong. Therefore, there must be an infinite number of primes!

@@

\begin{Proof}{}
The theorem of Mersenne can be proved by contradiction. We therefore assume that
there exists a composite natural number $ n $ (with real factorisation)  $ n=n_1
n_2 $, with the property that $ 2^n -1 $ is a prime number.

@@

Mersenne claimed that $2^{67}-1$ is a prime number. There is also a mathematical
history behind this claim: it first took over 200 years before \index{Lucas
  Edouard} Edouard Lucas (1842-1891) proved that this number is composite.
However, he argued indirectly and did not name any of the factors. Then Frank
Nelson showed in 1903 which factors make up this composite number: $$ 2^{67} -1
=147, 573, 952, 588, 676, 412, 927 = 193, 707, 721 \cdot 761, 838, 257, 287. $$
He admitted to having worked 20 years on the factorisation 
\index{factorisation} (expression as a product of factors) 
\footnote{%
Using CrypTool\index{CrypTool} v1.3 you can factorize numbers in the 
following way: menu {\bf Indiv. Procedures \textbackslash{} RSA Demonstration 
\textbackslash{} Factorisation of a Number}.
}
of this 21-digit decimal number!

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers 
\index{Mersenne!prime number} have the exponents 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217; ~ 4,253;\\
 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; ~ 132,049; \\
 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377; ~
6,972,593; \\13,466,917.
\end{array}
$$
Thus $39$ Mersenne prime numbers are currently known
\index{Prime number!Mersenne}\index{Mersenne!prime number}. 
For the first 38 Mersenne prime numbers we know that this list is complete.
The exponents until the 39th Mersenne prime number have not yet been checked
completely.

@@

The discoverer of M-38 received 50,000 USD from the EFF for discovering 
the first prime with more than 1 million decimal digits. 
The next prize from EFF for 100,000 USD is for a proven prime with more
than 10 million decimal digits.
%{\href{http://www.octocad.demon.co.uk/mersenne/prime.htm }{\tt http://www.octocad.demon.co.uk/mersenne/prime.htm}}.

Edouard Lucas (1842-1891) held the record for the longest prime number for over
70 years by proving that $2^{127}-1$ is prime. No new record is likely to last
that long.

@@

Up to now we have always looked for the prime factors in order to decide whether
a number is prime. However, even if the smallest prime factor is enormous, the
search takes too long. Factorising numbers using systematic computational
division or using the sieve of Eratosthenes (see below) is only feasible using
current computers for numbers with up to around $20$ digits in the decimal
system.

However, if we know something about the {\em construction} of the number in
question, there are extremely highly developed procedures that are much quicker.

@@

As early as in the 19th century, it was discovered that the $29$-digit number $$
F(7) = 2^{2^7} + 1 $$ is not prime. However, it was not until 1970 that
Morrison/Billhart managed to factorise it.
\begin{eqnarray*}
F(7) & = & 34,279,974,696,877,740,253,374,607,431,768,211,457 \\
& = & 59, 649, 589, 127, 497, 217 \cdot  574, 689, 200, 685,
129,054, 721.
\end{eqnarray*}
Many rapid prime number tests are based on the (little) Fermat theorem put
forward by Fermat in 1640 .
\index{Fermat!little theorem}
\begin{theorem}[``little'' Fermat]\label{thm-pz-fermat1}
Let $p$ be a prime number and $a$ be any whole number, then for all $a$ $$a^p
\equiv a \; {\rm mod} \; p.$$
This could also be formulated as follows: \\ Let $p$ be a prime number and $a$
be any whole number that is not a multiple of $p$ (also $a \not\equiv 0 \; {\rm
mod} \; p$), then $a^{p-1} \equiv 1 \; {\rm mod} \; p$.
\end{theorem}

@@

Comment: Numbers that have the property $$ 2^n \equiv 2 \;{\rm mod}\; n $$ but
are not prime are called \index{Pseudo prime number}\index{Number!pseudo prime} {\em pseudo prime numbers}. The first pseudo prime number (i.e. not a
prime) is $$ 341 = 11 \cdot 31. $$
There are numbers that pass the Fermat test with all bases and yet are not
prime: these numbers are called \index{Number!Carmichael} {\em
Carmichael numbers}. The first of these is $$ 561 = 3 \cdot 11 \cdot 17. $$
A stronger test is provided by \index{Miller} \index{Rabin} Miller/Rabin: it is
only passed by so-called {\em strong pseudo prime numbers}. Again, there are
strong pseudo prime numbers that are not primes, but this is much less often the
case than for (simple) pseudo prime numbers. The smallest strong pseudo prime
number base $2$ is $$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudo prime number up to $25 \cdot 10^9$, i.e. a number that passes the 
test and yet is not a prime number.

@@

\item {\em Mersenne numbers}  $f(n) = 2^n - 1$  \quad for $ n $ prime: \\
    As shown \hyperlink{MersenneNumbers01}{above}, this formula seems
    to deliver relatively large prime numbers but - as for $n=11$ 
    [$f(n)=2,047$] - it is repeatedly the case that the result even with
    prime exponents is \hyperlink{Mer-nums-not-always-prim}{not} prime.\\
    Today, all the Mersenne primes having less than 2,000,000 digits are 
    known (M-38) \\
    ({\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}).

\item $F(k,n) = k \cdot 2^n \pm 1$  for $ n $ prime and $ k $ small primes:\\
   For this generalisation of the Mersenne numbers there are (for small $k$)
   also extremely quick prime number tests (see \cite{Knuth1981}). This can
   be performed in practice using software such as the PROTHS software from
   Yves Gallot \\
   ({\href{http://www.prothsearch.net/index.html}
    {\tt http://www.prothsearch.net/index.html}}).

\vskip +15 pt
\hypertarget{FermatNumbers02}{}
\item {\em Fermat numbers}
\footnote{The Fermat prime numbers play a role in circle division.
As proven by Gauss\index{Gauss}, a regular $p$-edge
can only be constructed with the use of a pair of compasses and a ruler, when $p$ is a Fermat prime number.}
\index{Number!Fermat}\index{Fermat!number}  $F(n) = 2^{2^n} + 1$:\\
As mentioned \hyperlink{FermatNumbers01}{above}, Fermat wrote to Mersenne regarding his assumption, that all numbers of this type are primes.
Surprisingly he would have been able obtain a positive result using the negative
prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
F(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
F(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
F(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
F(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
F(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = 65,537 &  \mapsto {\rm ~prime}  \\
F(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = 4 ,294, 967, 297 = 641 \cdot
6,700,417 &  \mapsto {\rm ~NOT~prime} !
\end{array} 
$$

@@

    The Fermat number
    $$ F[2,145,351] = 2^{(2^{2,145,351})} + 1 $$ 
    is divisible by the prime
    $$ p = 3*2^{2,145,353} + 1 $$ \\
    This prime p is the largest known prime generalized Mersenne number. 
    It was discovered only a few days after Michael Angel discovered the 
    largest generalized Fermat number (\hyperlink{generalizedFermatprimes}
    {see below}).  This moved GIMPS' \index{GIMPS} first prime discovery,
    M-35 = M1398269, into 7th place.

    This work was done using NewPGen from Paul Jobling's, PRP from 
    George Woltman's, Proth from Yves Gallot's programs and also the
    Proth-Gallot group at St. Patrick's College, Dublin.

@@

\vskip +15 pt
\hypertarget{generalizedFermatprimes}{}
\item {\em Generalized Fermat numbers}\footnote{%
The base of this power is no longer restricted to 2 !}  $F(b,m) = b^{2^m} + 1$:
\index{Fermat!Fermat number!Generalized}  \\
    Generalized Fermat numbers are more numerous than Mersenne numbers at
    equal size and many of them are waiting to be discovered to fill the
    big gaps between the Mersenne primes already found or still undiscovered.
    Progress in number theory made it possible that numbers, where the
    representation is not limited to the base 2, can be tested at almost
    the same speed than a Mersenne number.

    Yves Gallot wrote the program Proth.exe to investigate generalized
    Fermat numbers.

@@

$e_9, \cdots, e_{17}$ are also composite, which means that this formula is not
particularly useful. Comment: However, what is special about these numbers
is that any pair of them does not have a common factor other than $1$. They are
therefore \index{Prime number!relative}\index{Relatively prim} {\em relatively prime}.

@@

Using the following Web site:
\vspace{-10pt}
\begin{itemize}
  \item[] \href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}
               {\tt http://www.math.Princeton.EDU/\~{}arbooker/nthprime.html}
\end{itemize}

you can easily determine $PI(x)$.

@@

\index{Eratosthenes!sieve}
\paragraph{Sieve of Eratosthenes}\mbox{}

@@

\begin{itemize}
  \item The German mathematician \index{Goldbach Christian} Christian Goldbach
      (1690-1764) conjectured: Every even natural number greater than 2 can
      be represented as the sum of two prime numbers.
      Computers have verified\footnote{%
It is generally accepted today, that the Goldbach Conjecture is true, 
i.~e.\ valid for all even natural numbers bigger than $2$.
In 1999, mathematician J\"org Richstein from the computer sciences 
institute at the University of Giessen, studied even numbers up to 
400 billion and found no contradictory example (see
\href{http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}
{\tt http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}).
\index{Richstein 1999}
Nevertheless, this does not provide us with general proof.\\
The fact is that despite all efforts, Goldbach's conjecture has to date
not been proven leads one to believe that since the pioneer work of
the Austrian mathematician Kurt G\"odel \index{G\""odel Kurt} is 
well-known, not every true mathematical theorem is provable (see
\href{http://www.mathematik.ch/mathematiker/goedel.html}
{http://www.mathematik.ch/mathematiker/goedel.html}).
Perhaps Goldbach's conjecture was correct, but in any case the proof
will never be found. Conversely, that will presumably also remain unproven.}
      the Goldbach conjecture for all even numbers up to
      $4*10^{14}$ but no general proof has yet been found\footnote{%
The English publisher {\em Faber} and the American publisher {\em Bloomsbury}
issued in 2000 the 1992 published book ``Uncle Petros and Goldbach's 
Conjecture'' by Apostolos Doxiadis.
It's the story of an old maths professor who fails to prove a more than
250 year old puzzle. To boost the sales figues the English and American
publishers have offered a prize of 1 million USD, if someone can prove
the conjecture -- which should be published by 2004 in a well-known
mathematical journal.\\
Surprisingly only British and American citizens are allowed to participate.\\
The theorem which has come closest so far to Goldbach's conjecture
was proved by Chen Jing-Run in 1966 in a way which is somewhat hard
to understand:
Each even integer greater than $2$ is the sum of one prime and of 
the product of two primes. E.g.: $20=5+3*5.$\\
Most of the research about the Goldbach conjecture is collected in 
the book: ``Goldbach Conjecture'', ed. Wang Yuan, 1984, World scientific
Series in Pure Maths, Vol. 4.\\
Especially this conjecture makes it clear, that even today we do not
have a complete understanding of the deeper connections between addition and
multiplication of natural numbers.}
      .
  \item The German \index{Riemann Bernhard} mathematician Bernhard Riemann
      (1826-1866) put forward a formula for the distribution of primes
      that would further improve the estimate. However, this has neither
      been proved nor contradicted so far.
\end{itemize}

@@

%already defined in elementaryNumberTheory.inc -> 2 davor
\item[Schneier1996]{2Schneier1996} \index{Schneier 1996}     
    Bruce Schneier, \\
    {\em Applied Cryptography}, Wiley and Sons, 2nd edition 1996.

@@

>>>>

The sequence of prime numbers starts with $$ 2,~ 3,~ 5,~ 7, ~ 11, ~ 13, ~
17, ~ 19, ~ 23, ~ 29, ~ 31, ~ 37, ~ 41, ~ 43, ~ 47, ~ 53, ~ 59, ~ 61,
~ 67, ~ 71, ~ 73, ~ 79, ~ 83, ~ 89, ~ 97, \cdots . $$
The first 100 numbers include precisely 25 prime numbers. After this,
the percentage of primes constantly decreases. Prime numbers can be
factorised in a uniquely {\em trivial} way: 
$$5 = 1 \cdot 5,\quad  17 = 1 \cdot 17, \quad 1,013 = 1 \cdot 1,013,  \quad
1,296,409 = 1 \cdot 1,296,409.$$
All numbers that have $2$ or more factors not equal 1 are called 
\index{Number!composite} {\em composite} numbers. 
These include $$ 4 = 2 \cdot 2, \quad 6 = 2\cdot 3 $$ as well
as numbers that {\em look like primes}, but are in fact composite:
$$ 91 = 7 \cdot 13, \quad 161=7 \cdot 23, \quad 767 =13 \cdot 59. $$

@@

\begin{theorem}\label{thm-pz-prod}
Each element $n$ of the natural numbers greater than $1$ can be written as the
product $n = p_1 \cdot p_2 \dots p_m$ of prime numbers. If two such
factorisations $$n =  p_1 \cdot p_2 \cdot \cdots \cdot p_m = p'_1 \cdot p'_2 \cdots
p'_{m'}$$ are given, then they can be reordered such that $\;m = m'\;$ and for
all $i$:  $\;p_i = p'_i$. \\
($n = p_1 \cdot p_2 \cdot \dots \cdot p_m$ are called the prime factors of n)

\end{theorem}

@@

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid%
\footnote{Euclid,
a Greek mathematician of 4th and 3rd century B.C. He worked at the
Egyptian academy of Alexandria and wrote ``The Elements'', the most well 
known systematically textbook of the Greek mathematics.} 
knew this in the third century B.C.
\begin{theorem}[Euclid\footnote{The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. The theorem has already been distinguished
and proven in Euclidís Elements (Book IX, s. 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as followed
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than 
any previously existing amount of prime numbers.
}]\label{thm-pz-euklid} % Ende der Fuﬂnote
% Fussnote VOR "]" in [Euclid], damit kein Leerraum vor der Fussnotennummer.
% Vorher stand da: ...(Euclid). BLANK Fussnote und das Blank stoerte.
% Nun steht die Fussnote direkt hinter "Euclid" und vor der ")".
% Eigentlich h‰tte ich sie gerne direkt hinter "(Euclid)", noch vor dem 
% automatisch gesetzten Punkt. 
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}
His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (proof by contradiction). He assumed that there is only a
finite number of primes and therefore a largest prime number. Based on this
assumption, he drew logical conclusions until he obtained an obvious
contradiction. This meant that something must be wrong. As there were no
mistakes in the chain of conclusions, it could only be the assumption that was
wrong. Therefore, there must be an infinite number of primes!

@@

\begin{Proof}{}
The theorem of Mersenne can be proved by contradiction. We therefore assume that
there exists a composite natural number $ n $ (with real factorisation)  
$ n=n_1 \cdot n_2 $, with the property that $ 2^n -1 $ is a prime number.

@@

Mersenne claimed that $2^{67}-1$ is a prime number. There is also a mathematical
history behind this claim: it first took over 200 years before \index{Lucas
  Edouard} Edouard Lucas (1842-1891) proved that this number is composite.
However, he argued indirectly and did not name any of the factors. Then Frank
Nelson showed in 1903 which factors make up this composite number: 
$$ 2^{67} -1
=147, 573, 952, 589, 676, 412, 927 = 193, 707, 721 \cdot 761, 838, 257, 287. $$
He admitted to having worked 20 years on the factorisation 
\index{factorisation} (expression as a product of prime factors) 
\footnote{%
Using CrypTool\index{CrypTool} v1.3 you can factorize numbers in the 
following way: menu {\bf Indiv. Procedures \textbackslash{} RSA Demonstration 
\textbackslash{} Factorisation of a Number}.
}
of this 21-digit decimal number!

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers 
\index{Mersenne!prime number} have the exponents 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217; ~ 4,253;\\
 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; ~ 132,049; \\
 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377; ~
6,972,593; \\13,466,917.
\end{array}
$$
Thus $39$ Mersenne prime numbers are currently known
\index{Prime number!Mersenne}\index{Mersenne!prime number}. 
For the first 38 Mersenne prime numbers we know that this list is complete.
The exponents until the 39th Mersenne prime number have not yet been checked
completely (see chapter 2.5 prime number tests\index{Prime number!test}).

@@

The discoverer of M-38 received 50,000 USD from the EFF for discovering 
the first prime with more than 1 million decimal digits. 
The next prize of 100,000 USD offered by EFF is for a proven prime with more
than 10 million decimal digits.
%{\href{http://www.octocad.demon.co.uk/mersenne/prime.htm }{\tt http://www.octocad.demon.co.uk/mersenne/prime.htm}}.

Edouard Lucas\index{Lucas Edouard} (1842-1891) held the record for the
longest prime number for over 70 years by proving that $2^{127}-1$ is prime.
No new record is likely to last that long.

@@

Up to now we have looked for the prime factors in order to decide whether
a number is prime. However, even if the smallest prime factor is enormous, the
search takes too long. Factorising numbers using systematic computational
division or using the \hyperlink{SieveEratosthenes01}{sieve of Eratosthenes} 
\index{Eratosthenes!sieve} is only feasible using current computers for
numbers with up to around $20$ digits in the decimal system.
The biggest number factorized into its 2 almost equal prime factors 
has 158 digits (see \hyperlink{C158-chap3}{chapter 3.11.4}).

However, if we know something about the {\em construction} of the number in
question, there are extremely highly developed procedures that are much quicker.
These procedures can determine the primality attribute of a number, but they
cannot determine the prime factors of the numbers, if they are compound.

@@

As early as in the 19th century, it was discovered that the $29$-digit number $$
F(7) = 2^{2^7} + 1 $$ is not prime. However, it was not until 1970 that
Morrison/Billhart managed to factorise it.
\begin{eqnarray*}
F(7) & = & 34,279,974,696,877,740,253,374,607,431,768,211,457 \\
& = & 59, 649, 589, 127, 497, 217 \cdot  574, 689, 200, 685,
129,054, 721.
\end{eqnarray*}

\\
Despite Fermat was wrong with this supposition, he is the originator of
an important theorem in this area: Many rapid prime number tests are 
based on the (little) Fermat theorem put forward by Fermat in 1640
(\hyperlink{KleinerSatzFermat-chap3}{see chapter 3.8.3}).

\hypertarget{KleinerSatzFermat-chap2}{}
\index{Fermat!little theorem}
\begin{theorem}[``little'' Fermat]\label{thm-pz-fermat1}
Let $p$ be a prime number and $a$ be any whole number, then for all $a$ $$a^p
\equiv a \; {\rm mod} \; p.$$
This could also be formulated as follows: \\ Let $p$ be a prime number and $a$
be any whole number that is not a multiple of $p$ (also $a \not\equiv 0 \; {\rm
mod} \; p$), then $a^{p-1} \equiv 1 \; {\rm mod} \; p$.
\end{theorem}

@@

Comment: Numbers n that have the property $$ 2^n \equiv 2 \;{\rm mod}\; n $$ but
are not prime are called \index{Prime number!pseudo prime}\index{Number!pseudo prime} {\em pseudo prime numbers}. The first pseudo prime number (i.e. not a
prime) is $$ 341 = 11 \cdot 31. $$
There are numbers that pass the Fermat test with all bases and yet are not
prime: these numbers are called \index{Number!Carmichael} {\em
Carmichael numbers}. The first of these is $$ 561 = 3 \cdot 11 \cdot 17. $$

A stronger test is provided by\index{Miller}\index{Rabin} Miller/Rabin: it is
only passed by so-called {\em strong pseudo prime numbers}. Again, there are
strong pseudo prime numbers that are not primes, but this is much less often the
case than for (simple) pseudo prime numbers. The smallest strong pseudo prime
number base $2$ is $$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudo prime number up to $25 \cdot 10^9$, i.e. a number that passes the 
test and yet is not a prime number.

@@

\item {\em Mersenne numbers}  $f(n) = 2^n - 1$  \quad for $ n $ prime: \\
    As shown \hyperlink{MersenneNumbers01}{above}, this formula seems
    to deliver relatively large prime numbers but - as for $n=11$ 
    [$f(n)=2,047$] - it is repeatedly the case that the result even with
    prime exponents is \hyperlink{Mer-nums-not-always-prim}{not} prime.\\
    Today, all the Mersenne primes having less than around 2,000,000 digits are 
    known (M-38\index{M-38}) \\
    ({\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}).

\item $F(k,n) = k \cdot 2^n \pm 1$  for $ n $ prime and $ k $ small primes:\\
   For this generalisation of the Mersenne 
   numbers\index{Mersenne!number!generalized}
   there are (for small $k$)
   also extremely quick prime number tests (see \cite{Knuth1981}). This can
   be performed in practice using software such as the Proths software from
   Yves Gallot\index{Gallot Yves} \\
   ({\href{http://www.prothsearch.net/index.html}
    {\tt http://www.prothsearch.net/index.html}}).

\vskip +15 pt
\hypertarget{FermatNumbers02}{}
\item {\em Fermat numbers}\footnote{The Fermat prime numbers play a role in circle division.
As proven by Gauss\index{Gauss}, a regular $p$-edge
can only be constructed with the use of a pair of compasses and a ruler, when $p$ is a Fermat prime number.}
\index{Number!Fermat}\index{Fermat!number}  $F(n) = 2^{2^n} + 1$:\\
As mentioned \hyperlink{FermatNumbers01}{above}, Fermat wrote to Mersenne regarding his assumption, that all numbers of this type are primes.
Surprisingly he would have been able obtain a positive result using the negative
prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
F(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
F(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
F(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
F(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
F(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = 65,537 &  \mapsto {\rm ~prime}  \\
F(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = 4 ,294, 967, 297 = 641 \cdot
6,700,417 &  \mapsto {\rm ~NOT~prime} !
\end{array} 
$$

@@

    The Fermat number
    $$ F[2,145,351] = 2^{(2^{2,145,351})} + 1 $$ 
    is divisible by the prime
    $$ p = 3*2^{2,145,353} + 1 $$ \\
    This prime p is the largest known prime generalized Mersenne 
    number\index{Mersenne!number!generalized}. 
    It was discovered only a few days after Michael Angel discovered the 
    largest generalized Fermat number (\hyperlink{generalizedFermatprimes}
    {see below}).  This moved GIMPS' \index{GIMPS} first prime discovery,
    M-35 = M1398269, into 7th place.

    This work was done using NewPGen from Paul Jobling's, PRP from 
    George Woltman's, Proth from Yves Gallot's programs\index{Gallot Yves} and also the
    Proth-Gallot group at St. Patrick's College, Dublin.

@@

\vskip +15 pt
\hypertarget{generalizedFermatprimes}{}
\item {\em Generalized Fermat numbers}\footnote{%
The base of this power is no longer restricted to 2 !}  $F(b,m) = b^{2^m} + 1$:
\index{Fermat!number!generalized}  \\
    Generalized Fermat numbers are more numerous than Mersenne numbers at
    equal size and many of them are waiting to be discovered to fill the
    big gaps between the Mersenne primes already found or still undiscovered.
    Progress in number theory made it possible that numbers, where the
    representation is not limited to the base 2, can be tested at almost
    the same speed than a Mersenne number.

    Yves Gallot\index{Gallot Yves} wrote the program Proth.exe to investigate generalized
    Fermat numbers.

@@

$e_9, \cdots, e_{17}$ are also composite, which means that this formula is not
particularly useful. Comment: However, what is special about these numbers
is that any pair of them does not have a common factor other than $1$. They are
therefore\index{Prime number!relative}\index{Relatively prim} {\em relatively prime}.

@@

Using the following Web site:
\vspace{-10pt}
\begin{itemize}
  \item[] \href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}
               {\tt http://www.math.Princeton.EDU/\~{}arbooker/nthprime.html}
\end{itemize}
you can easily determine $PI(x)$.

@@

%----------------------------------------
\index{Eratosthenes!sieve}
\paragraph{Sieve of Eratosthenes}\mbox{}
\hypertarget{SieveEratosthenes01}{}

@@

\begin{itemize}
  \item The German mathematician \index{Goldbach Christian} Christian Goldbach
      (1690-1764) conjectured: Every even natural number greater than 2 can
      be represented as the sum of two prime numbers.
      Computers have verified\footnote{%
It is generally accepted today, that the Goldbach Conjecture is true, 
i.~e.\ valid for all even natural numbers bigger than $2$.
In 1999, mathematician J\"org Richstein from the computer sciences 
institute at the University of Giessen, studied even numbers up to 
400 billion and found no contradictory example (see
\href{http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}
{\tt http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}).
\index{Richstein 1999}
Nevertheless, this does not provide us with general proof.\\
The fact is that despite all efforts, Goldbach's conjecture has to date
not been proven. This leads one to believe that since the pioneer work of
the Austrian mathematician Kurt G\"odel \index{G\""odel Kurt} is 
well-known, not every true mathematical theorem is provable (see
\href{http://www.mathematik.ch/mathematiker/goedel.html}
{http://www.mathematik.ch/mathematiker/goedel.html}).
Perhaps Goldbach's conjecture was correct, but in any case the proof
will never be found. Conversely, that will presumably also remain unproven.}
      the Goldbach conjecture for all even numbers up to
      $4*10^{14}$ but no general proof has yet been found\footnote{%
The English publisher {\em Faber} and the American publisher {\em Bloomsbury}
issued in 2000 the 1992 published book ``Uncle Petros and Goldbach's 
Conjecture'' by Apostolos Doxiadis.
It's the story of an old maths professor who fails to prove a more than
250 year old puzzle. To boost the sales figures the English and American
publishers have offered a prize of 1 million USD, if someone can prove
the conjecture -- which should be published by 2004 in a well-known
mathematical journal.\\
Surprisingly only British and American citizens are allowed to participate.\\
The theorem which has come closest so far to Goldbach's conjecture
was proved by Chen Jing-Run in 1966 in a way which is somewhat hard
to understand:
Each even integer greater than $2$ is the sum of one prime and of 
the product of two primes. E.g.: $20=5+3*5.$\\
Most of the research about the Goldbach conjecture is collected in 
the book: ``Goldbach Conjecture'', ed. Wang Yuan, 1984, World scientific
Series in Pure Maths, Vol. 4.\\
Especially this conjecture makes it clear, that even today we do not
have a complete understanding of the deeper connections between addition and
multiplication of natural numbers.}.

  \item The German \index{Riemann Bernhard} mathematician Bernhard Riemann
      (1826-1866) put forward a formula for the distribution of primes
      that would further improve the estimate. However, this has neither
      been proved nor contradicted so far.
\end{itemize}

@@

%already defined in elementaryNumberTheory.inc -> 2 davor
\item[Schneier1996]{2Schneier1996} \index{Schneier 1996}     
    Bruce Schneier, \\
    {\em Applied Cryptography, Protocols, Algorithms, and Source Code in C},\\
    Wiley and Sons, 2nd edition 1996.

@@

\item The best overview about prime numbers is offered from my point of view 
      by ~``The Prime Pages'' from Chris Caldwell.
      \index{Chris Caldwell} \\
      \href{http://www.utm.edu/research/primes}
       {\tt http://www.utm.edu/research/primes }
